SFJ: An implementation of Semantic Featherweight Java

### 2020-04-20 conference/camera ready/paper.tex
working on background

\title{SFJ: An implementation of Semantic Featherweight Java
    \thanks{Research supported by the UK EPSRC grant EP/K034413/1, ``From Data Types to Session Types: A Basis for Concurrency and Distribution (ABCD)'', and by the EU HORIZON 2020 MSCA RISE project 778233 ``BehAPI: Behavioural Application Program Interfaces''.}
}
\titlerunning{SFJ}

@@

    An orthogonal subtyping question, typical of object-oriented languages, is the \emph{nominal} and the \emph{structural} subtyping.
    Dardha \emph{et al.} \cite{Dardha2013,Dardha2017} defined boolean types and semantic subtyping for Featherweight Java (FJ) and integrated both nominal and structural subtyping, thus exploiting the benefits of both approaches.
    However, these benefits were illustrated only at a theoretical level, but not exploited practically.

@@

\section{Introduction}
There are two approaches to defining subtyping relations: the \emph{syntactic} and the \emph{semantic} approach.
Syntactic subtyping is more mainstream in programming languages and is defined by means of a set of formal deductive subtyping rules.
Semantic subtyping is less know:  one defines a formal model of the  language and an interpretation of types as subsets of this model, ultimately, these will be subsets of values in the language. Then, subtyping is defined as set inclusion of these subsets denoting types.
%
Orthogonally, for object-oriented languages there are two approaches to defining subtyping relations: the \emph{nominal} and the \emph{structural} approach.
Nominal subtyping is based on {declarations} by the developer and is \emph{name}-based: $A$ is a subtype of $B$ if and only if it is declared to be so, that is if the class $A$ extends (or implements) the class (or interface) $B$.
Structural subtyping instead is based on the \emph{structure} of a class, its fields and methods: a class $A$ is a subtype of a class $B$ if and only if the fields and methods of $A$ are a superset of the fields and methods of $B$, and their types in $A$ are subtypes of the types in $B$.
It is not surprising that the nominal subtyping aligns well with the syntactic approach, and in fact it is more well-known than the structural subtyping, which in turn aligns well with the semantic approach.

Dardha \emph{et al.} \cite{Dardha2013,Dardha2017} define {boolean types} and semantic subtyping for \emph{Featherweight Java} (FJ) \cite{featherweight}.
This approach allows for the integration of both nominal and structural subtyping in FJ, bringing in higher guarantees of program correctness, flexibility and compactness in programming.
Unfortunately, these benefits were only presented at a theoretical level and not exploited practically, due to the lack of an implementation of the language and its types and type system.

In this paper, we present SFJ---\emph{Semantic Featherweight Java} \autoref{sec:design}, an implementation of FJ with boolean types and semantic subtyping.
In SFJ the developer has a larger and more expressive set of types, by using boolean connectives \textbf{and}, \textbf{not}, \textbf{or}, with the expected set-theoretic interpretation. On the other hand, this added expressivity does not introduce added complexity, but rather the opposite is true, as the developer has an easier, more compact and more elegant way of programming.
SFJ integrates both structural and nominal subtyping, and the developer can choose which one to use.
Finally, as discussed in Dardha \emph{et al.} \cite[\S 8.4]{Dardha2017}, thanks to semantic subtyping, we can easily encode in SFJ standard programming constructs and features of Java, such as lists, or overloading classes via multimethods \cite{BC97}, which are missing in FJ, thus making SFJ a more complete language closer to Java.

@@

    Now, suppose that our class hierarchy is such that \emph{Polygon} is the parent class and all other geometric figure classes extend \emph{Polygon}, which is how one would naturally define the set of polygons.
    If we think of the class hierarchy as given and part of legacy code, which cannot be changed, then again a natural way to implement this in Java is by defining the method \emph{diagonal} in the class \emph{Polygon} and using an \textbf{instanceof}, for e.g., inside a \textbf{try-catch} which throws an exception at runtime, if the argument passed to the method is a triangle.

    We propose a more elegant solution, by combining boolean types and semantic subtyping, where only static typechecking is required and implement this in SFJ \cite{UD20}: it is enough to define a method \emph{diagonal} that has an argument of type \emph{Polygon} \textbf{and not} \emph{Triangle}, thus allowing the typechecker to check at compile time the restrictions on types:
    $$
        \begin{array}{l}
            \synclass Polygon \ \{\ldots\}
            \\
            \syndecl{Triangle}{Polygon} {\ldots}
            \\
            \syndecl{Square}{Polygon} {\ldots}
            \\
            \syndecl{Rhombus}{Polygon} {\ldots}
            \\
            \hspace{2.5cm} \ldots
            \\
            \synclass Diagonal \ \{
            \\
            %\qquad Diagonal ( )\   \{\ super();\ \}
            \hspace{2.5cm} \ldots \\
            \qquad \mathbf{double}\ diagonal((\textit{Polygon} \textbf{ and not }  \textit{Triangle}) \ shape)\{\ldots\}
            \\
            \}
        \end{array}
    $$
    We can now call \emph{diagonal} on an argument of type \emph{Polygon}: if the polygon is \textbf{not} a \emph{Triangle}, then the method returns the length of its longest diagonal; otherwise, if the polygon is a triangle, then there will be a type error at compile time.
\end{example}

The technical developments behind semantic subtyping and its properties are complex, however, they are completely transparent to the developer.
Due to space limit we will remove most theoretical details from this submission, as they are not necessary to understand the design and implementation of SFJ, which is the focus of this paper. We refer the reader to the relevant work \cite{Dardha2013,Dardha2017} and we will include a more complete background on semantic subtyping in a possible full paper.

\section{SFJ: Design and Implementation}
\label{sec:design}
The syntax of types is given by the following grammar \cite{Dardha2013,Dardha2017}:

$$
    \begin{array}{llll}
        \tau   & ::=                                       & \alpha \ |\ \mu
               & \mbox{{\em Type term}}
        \\
        \alpha & ::=                                       & \zero \ |\ \btypes \ |\ [\wt{l:\tau}] \ |\ \alpha \ \bf{and}\ \alpha \ |\ \bf{not}\ \alpha
        \qquad
               & \mbox{{\em Object type} ($\alpha $-type)}
        \\
        \mu    & ::=                                       & \alpha \to \alpha \ |\ \mu \ \bf{and}\ \mu \ |\ \bf{not}\ \mu
               & \mbox{{\em Method type} ($\mu $-type)}
    \end{array}
$$

$\alpha$-types are used to type fields and $\mu$-types are used to type methods.
Type $\zero$ is the empty type.
Type $\btypes$ denotes the \emph{basic} types, such as integers, booleans, etc.
    {Record} types $[\wt{l:\tau}]$, where $\wt l$ is a sequence of disjoint labels are used to type objects.
Arrow types $\alpha \to \alpha$ are used to type methods.
The boolean types using $\bf{and}$ and $\bf{not}$ have their expected set-theoretic meanings, and $\bf{or}$ is obtained by their combination.\\
The syntax of terms is given by the following grammar and is based on the standard syntax of terms in FJ \cite{featherweight,Dardha2013,Dardha2017}:
\begin{align*}
     & \mbox{\textit{Class declaration}}  & L \; ::= \; & \syndecl{C}{C}{\wt{\alpha \ a};\ K; \ \wt{M}\ }                         \\
     & \mbox{\textit{Constructor}}        & K \;::=\;   & C\ (\wt{\alpha\ x})\ \{\ \synsuper(\wt{x});\ \wt{\synthis.a}=\wt{x}; \} \\
     & \mbox{\textit{Method declaration}} & M \; ::= \; & \alpha \ m\ (\alpha \ x)\ \{\ \synret e; \}                             \\
     & \mbox{\textit{Expressions}}        & e \; ::=\;  & x\ |\  c\ |\ e.a\ |\ e.m(e) \ | \ \synnew C(\wt{e})
\end{align*}
Since we want to use types $\tau$ in practice in SFJ, we restrict them to finite trees whose leaves are constants with no cycles.
For e.g., a recursive type $\alpha = [a : \alpha]$ denotes an infinite program tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)), hence we avoid it as it is inhabitable.
%Similarly the types $\alpha = [b: \beta]$, $\beta = [a = \alpha]$ would also be impossible to instantiate.
When processing the abstract syntax tree (AST) of a program in SFJ, we forbid fields to have the same type as the type of the class they are defined in.
Secondly, we mark any classes containing fields typed with only basic types as \emph{resolved} otherwise, as \emph{unresolved}.
This is captured by the following algorithm, which checks if the type definitions in the program are valid.
\begin{verbatim}
do
    boolean resolutionOccured = false
    for class that is unresolved:
        boolean resolved = true
        for field in class that contains a class type:
            if class type is not resolved
                resolved = false
        if resolved is true:
            set class to be resolved
            resolutionOccured = true
while resolutionOccured is true

@@

The subtyping algorithm finds all nominal and structural subtypes.
It finds all nominal subtypes as they inherit all fields and methods of their super class, so are always guaranteed to be a superset.
Similarly, given all pairs of types, the algorithm checks if one is a superset of the other, i.e.,\ they share common fields and methods, and if so, they are related by structural subtyping.
For e.g., type $empty = []$, will have all classes as structural subtypes.
%The above algorithm will without differentiating, find all nominal and structural subtypes as we always consider all other classes in the program to be potential subtype candidates.

@@

To conclude, in this paper we presented the design and implementation of SFJ, an extension of Featherweight Java with boolean types and semantic subtyping, which allow for both structural and nominal subtyping in FJ as well as restore standard Java constructs for e.g., lists and features for e.g., overloading which were not present in FJ, making SFJ a more complete language.

%% References
\newpage
\bibliographystyle{splncs04}
\bibliography{paper}

\end{document}

>>>>

\title{SFJ: An implementation of Semantic Featherweight Java
    \thanks{Supported by the UK EPSRC grant EP/K034413/1, ``From Data Types to Session Types: A Basis for Concurrency and Distribution'' (ABCD), and by the EU HORIZON 2020 MSCA RISE project 778233
    ``Behavioural Application Program Interfaces'' (BehAPI).}
}
\titlerunning{SFJ}

@@

    An orthogonal subtyping question, typical of object-oriented languages, is the \emph{nominal} versus the \emph{structural} subtyping.
    Dardha \emph{et al.} \cite{Dardha2013,Dardha2017} defined boolean types and semantic subtyping for Featherweight Java (FJ) and integrated both nominal and structural subtyping, thus exploiting the benefits of both approaches.
    However, these benefits were illustrated only at a theoretical level, but not exploited practically.

@@

\section{Introduction}
There are two approaches to defining subtyping relations: the \emph{syntactic} and the \emph{semantic} approach.
Syntactic subtyping \cite{Liskov1994} is more mainstream in programming languages and is defined by means of a set of formal deductive subtyping rules.
Semantic subtyping \cite{Castagna2005,Cas05,Frisch2008} is more recent and less known:  one defines a formal model of the language and an interpretation of types as subsets of this model. Then, subtyping is defined as set inclusion of subsets denoting types.

Orthogonally, for object-oriented languages there are two approaches to defining subtyping relations: the \emph{nominal} and the \emph{structural} approach \cite{Integrating,Useful}.
Nominal subtyping is based on {declarations} by the developer and is \emph{name}-based: ``$A$ is a subtype of $B$ if and only if it is declared to be so, that is if the class $A$ extends (or implements) the class (or interface) $B$''.
Structural subtyping instead is based on the \emph{structure} of a class, its fields and methods: ``a class $A$ is a subtype of a class $B$ if and only if the fields and methods of $A$ are a superset of the fields and methods of $B$, and their types in $A$ are subtypes of the types in $B$''.
Intutively, extending a class with more fields and methods results in a smaller set of inhabitants for that class. For example, the set of objects of a class \emph{Student} is smaller than the set of objects of a class \emph{Person} as each Student is a Person, but not the other way around. Hence, \emph{Student} is a structural subtype of \emph{Person}, even if it is not declared to be so.

Dardha \emph{et al.} \cite{Dardha2013,Dardha2017} define boolean types---based on set-theoretic operations such as \textbf{and}, \textbf{not}, \textbf{or}---and semantic subtyping for \emph{Featherweight Java} (FJ) \cite{featherweight}.
This approach allows for the integration of both nominal and structural subtyping in FJ, bringing in higher guarantees of program correctness, flexibility and compactness in program writing.
Unfortunately, these benefits were only presented at a theoretical level and not exploited practically, due to the lack of an implementation of the language and its types and type system.

In this paper, we present SFJ---\emph{Semantic Featherweight Java} \autoref{sec:design}, an implementation of FJ with boolean types and semantic subtyping.
In SFJ the developer has a larger and more expressive set of types, by using boolean connectives \textbf{and}, \textbf{not}, \textbf{or}, with the expected set-theoretic interpretation. On the other hand, this added expressivity does not add complexity. Rather the opposite is true, as the developer has an easier, more compact and elegant way of programming.
SFJ integrates both structural and nominal subtyping, and the developer can choose which one to use.
Finally, as discussed in Dardha \emph{et al.} \cite[\S 8.4]{Dardha2017}, thanks to semantic subtyping, we can easily encode in SFJ standard programming constructs and features of the full Java language, such as lists, or overloading classes via multimethods \cite{BC97}, which are missing in FJ, thus making SFJ a more complete language closer to Java.

@@

    Now, suppose our class hierarchy is such that \emph{Polygon} is the parent class and all other geometric figures extend \emph{Polygon}, which is how one would naturally define the set of polygons.
    Suppose the class hierarchy is given and is part of legacy code, which cannot be changed. Then again, a natural way to implement this in Java is by defining the method \emph{diagonal} in the class \emph{Polygon} and using an \textbf{instanceof}, for example, inside a \textbf{try-catch} construct. Then, an exception would be thrown at runtime, if the argument passed to the method is a triangle.

    We propose a more elegant solution, by combining boolean types and semantic subtyping, where only static typechecking is required and we implement this in SFJ \cite{UD20}: it is enough to define a method \emph{diagonal} that has an argument of type \emph{Polygon} \textbf{and not} \emph{Triangle}, thus allowing the typechecker to check at compile time the restrictions on types:
    $$
        \begin{array}{l}
            \synclass Polygon \ \{\ldots\}
            \\
            \syndecl{Triangle}{Polygon} {\ldots}
            \\
            \syndecl{Square}{Polygon} {\ldots}
            \\
            \syndecl{Rhombus}{Polygon} {\ldots}
            \\
            \hspace{2.5cm} \ldots
            \\
            \synclass Diagonal \ \{
            \\
            %\qquad Diagonal ( )\   \{\ super();\ \}
            \hspace{2.5cm} \ldots \\
            \qquad \mathbf{double}\ diagonal((\textit{Polygon} \textbf{ and not }  \textit{Triangle}) \ shape)\{\ldots\}
            \\
            \}
        \end{array}
    $$
    We can now call \emph{diagonal} on an argument of type \emph{Polygon}: if the polygon is \textbf{not} a \emph{Triangle}, then the method returns the length of its longest diagonal; otherwise, if the polygon is a triangle, then there will be a type error at compile time.
\end{example}

%The technical developments behind semantic subtyping and its properties are complex, however, they are completely transparent to the developer.
%Due to space limit we will remove most theoretical details from this submission, as they are not necessary to understand the design and implementation of SFJ, which is the focus of this paper. We refer the reader to the relevant work \cite{Dardha2013,Dardha2017} and we will include a more complete background on semantic subtyping in a possible full paper.

The rest of the paper is organised as follows:
\autoref{sec:background} details the technical developments behind semantic subtyping and its properties.
\autoref{sec:design} details the design and implementation of SFJ.
\autoref{sec:conclusion} concludes the paper.

\input{background.tex}

\section{SFJ: Design and Implementation}
\label{sec:design}
Since we want to use types $\tau$ in practice in SFJ, we restrict them to finite trees whose leaves are constants with no cycles.
For example, a recursive type $\alpha = [a : \alpha]$ denotes an infinite program tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)), hence we avoid it as it is inhabitable.
%Similarly the types $\alpha = [b: \beta]$, $\beta = [a = \alpha]$ would also be impossible to instantiate.
When processing the abstract syntax tree (AST) of a program in SFJ, we forbid fields to have the same type as the type of the class they are defined in.
Secondly, we mark any classes containing fields typed with only basic types as \emph{resolved} otherwise, as \emph{unresolved}.
This is captured by the following algorithm, which checks if the type definitions in the program are valid.
\begin{verbatim}
do
    boolean resolutionOccured = false
    for class that is unresolved:
        boolean resolved = true
        for field in class that contains a class type:
            if class type is not resolved
                resolved = false
        if resolved is true:
            set class to be resolved
            resolutionOccured = true
while resolutionOccured is true

@@

The subtyping algorithm finds all nominal and structural subtypes.
It finds all nominal subtypes as they inherit all fields and methods of their super class, so are always guaranteed to be a superset.
Similarly, given all pairs of types, the algorithm checks if one is a superset of the other, i.e.,\ they share common fields and methods, and if so, they are related by structural subtyping.
For example, type $empty = []$, will have all classes as structural subtypes.
%The above algorithm will without differentiating, find all nominal and structural subtypes as we always consider all other classes in the program to be potential subtype candidates.

@@

To conclude, in this paper we presented the design and implementation of SFJ, an extension of Featherweight Java with boolean types and semantic subtyping, which allow for both structural and nominal subtyping in FJ as well as restore standard Java constructs for example, lists and features for example, overloading which were not present in FJ, making SFJ a more complete language.

%% References
\newpage
\bibliographystyle{splncs04}
\bibliography{paper}
\end{document}
